Peaks and valleys of prolatin-related gene expression during pigeon parental care stages

### 2019-09-06 docs/characterization_manuscript.Rmd
more abstract
now with char figs

---
title: Peaks and valleys of prolatin-related gene expression during pigeon parental care stages
author:
  - name: Rayna M Harris
    email: rmharris@ucdavis
    affiliation: UC Davis
    corresponding: "rmharris@ucdavis"
  - name: Suzanne Austin
    affiliation: UC Davis
  - name: Andrew Lang
    affiliation: UNH
  - name: Matthew MacManes
    affiliation: UNH
  - name: Rebecca M Calisi
    affiliation: UC Davis           
address:
  - code: UC Davis
    address: UC Davis, Davis, CA
  - code: UNH
    address: UNH, smalltown, NH
abstract: |
  The goal of this research was to provide the most comprehensive gene expression profile of gene expression activity in the HPG axis of male and female rock doves to provide a deeper undertand of how the reproductive axis response to typical behavioral transitions associated with parental care. Here we used RNA sequencing to measure gene activity at 8 stages of the parental care cyle from nest building, to egg incubation, and through nestling care. Non-parental groups were added as controls, but gene expressions differences were assesed between and across all time points. The three tissue have unique signatures, but within each tissue we find minial sex differences. The pitutary samples displacy the msot plasticity in gene expression, with many changes in gene expression mirroring the typical rise and fall of circulating prolatin that peaks when chicks hatch. Our analysis provides new insight into how suites of genes respond in concert to the demands of offspring care.
author_summary: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. 
bibliography: mybibfile.bib
output: rticles::plos_article
csl: plos.csl
---

@@

# install.packages(c('tinytex', 'rmarkdown'))
# tinytex::install_tinytex()
```

@@

Rock doves are socially monogamous and offer bi-parental care, making inter- and intra-sexual comparisons possible. Birds offer two important behavioral transition points into parental care: the incubation of eggs and the caring for chicks. This produces two unique opportunities to study how the brain transitions into two different suites of parental care behaviors. Additionally, rock doves exhibit a parental care strategy analogous to mammals in that they, too, ‘lactate’ to feed their young (Gillespie et al. 2011, 2012). This lactation, unlike simple regurgitation of food, consists of the production and sloughing off of skin cells inside the crop sac of females and males, creating a protein-rich milk-like substance on which they rear their chicks. Many functional similarities between rock dove and mammalian lactation exist concerning the mediation of this event by the hormone prolactin (Dumont 1965). Additionally, like mammalian milk, rock dove milk delivers essential immunoglobulins and nutritional benefits to young, aiding in their immune function and development of microbiota @Gillespie2012. Thus, because rock doves incubate eggs and exhibit mammalian-like mediation and function of lactation for young, they have the potential to serve as a powerful theoretical bridge to understand the neurobiology of both avian and mammalian transitions into parental care.

# Materials and Methods and Results

```{r, out.width = "360px", echo = F}
knitr::include_graphics("../figures/images/samplingtimepoints.png")
```

## Characterize changes in neural-gene expression during parental care transitions

Our working hypothesis is that distinct changes in transcription occur in the brain at the anticipation of, during, and in response to two different types of parental care: incubation behavior and hatchling care.

```{r, message=FALSE, warning=FALSE, cache=TRUE, echo = F, fig.width= 5}
library(cowplot)
ggdraw() + draw_image("../figures/characterization/linegraphs-1.png") 
```

**Fig 2. The magnitude of gene expression changes between each parental transition**.
4-5K genes are differentially expressed beteen control birds and their nest buliding conspecifics in all tissues except the male gonad. 500 - 1000 genes are differentially expressed in the female pituitary from mid-late incubation as well as in the female hypothalamus and pituitary from hatch to nestingly care day 5.

>>>>

---
title: Peaks and valleys of prolatin-related gene expression during pigeon parental care stages
author:
  - name: Rayna M Harris
    email: rmharris@ucdavis
    affiliation: UC Davis
    corresponding: "rmharris@ucdavis"
  - name: Suzanne Austin
    affiliation: UC Davis
  - name: Andrew Lang
    affiliation: UNH
  - name: Matthew MacManes
    affiliation: UNH
  - name: Rebecca M Calisi
    affiliation: UC Davis           
address:
  - code: UC Davis
    address: UC Davis, Davis, CA
  - code: UNH
    address: UNH, smalltown, NH
abstract: |
  The goal of this research was to provide the most comprehensive gene expression profile of gene expression activity in the HPG axis of male and female rock doves to provide a deeper undertand of how the reproductive axis response to typical behavioral transitions associated with parental care. Here we used RNA sequencing to measure gene activity at 8 stages of the parental care cyle from nest building, to egg incubation, and through nestling care. Non-parental groups were added as controls, but gene expressions differences were assesed between and across all time points. The three tissue have unique signatures, but within each tissue we find minial sex differences. The pitutary samples displacy the msot plasticity in gene expression, with many changes in gene expression mirroring the typical rise and fall of circulating prolatin that peaks when chicks hatch. Our analysis provides new insight into how suites of genes respond in concert to the demands of offspring care. This data can be used to develop and test hypotheses about the mechanism regulating parental care behavior. 
author_summary: |
  Lorem ipsum dolor sit amet, consectetur adipiscing elit. 
bibliography: mybibfile.bib
output: rticles::plos_article
csl: plos.csl
---

@@

# install.packages(c('tinytex', 'rmarkdown'))
# tinytex::install_tinytex()

library(cowplot)
```

@@

Rock doves are socially monogamous and offer bi-parental care, making inter- and intra-sexual comparisons possible. Birds offer two important behavioral transition points into parental care: the incubation of eggs and the caring for chicks. This produces two unique opportunities to study how the brain transitions into two different suites of parental care behaviors. Additionally, rock doves exhibit a parental care strategy analogous to mammals in that they, too, ‘lactate’ to feed their young (Gillespie et al. 2011, 2012). This lactation, unlike simple regurgitation of food, consists of the production and sloughing off of skin cells inside the crop sac of females and males, creating a protein-rich milk-like substance on which they rear their chicks. Many functional similarities between rock dove and mammalian lactation exist concerning the mediation of this event by the hormone prolactin (Dumont 1965). Additionally, like mammalian milk, rock dove milk delivers essential immunoglobulins and nutritional benefits to young, aiding in their immune function and development of microbiota @Gillespie2012. Thus, because rock doves incubate eggs and exhibit mammalian-like mediation and function of lactation for young, they have the potential to serve as a powerful theoretical bridge to understand the neurobiology of both avian and mammalian transitions into parental care. Our working hypothesis is that distinct changes in transcription occur in the brain at the anticipation of, during, and in response to two different types of parental care: incubation behavior and hatchling care.

# Materials and Methods
```{r, out.width = "360px", echo = F}
knitr::include_graphics("../figures/images/samplingtimepoints.png")
```

# Results

## Table 1: comparison of total DEGs determined by edgeR and DESeq2

## Fig 1: Total number of gene expression differences between each stage

```{r, message=FALSE, warning=FALSE, cache=TRUE, echo = F, fig.width= 5}
ggdraw() + draw_image("../figures/characterization/linegraphs-1.png") 
```

**Fig x. The magnitude of gene expression changes between each parental transition**.
4-5K genes are differentially expressed beteen control birds and their nest buliding conspecifics in all tissues except the male gonad. 500 - 1000 genes are differentially expressed in the female pituitary from mid-late incubation as well as in the female hypothalamus and pituitary from hatch to nestingly care day 5.

## Fig 2: Ability to use gene expression to distingush groups

```{r, message=FALSE, warning=FALSE, cache=TRUE, echo = F, fig.width= 5}
ggdraw() + draw_image("../figures/sexes/DoveParentsRNAseq_model2.png") 
```

## Fig 3: Prolactin expression in PIT (but not HYP or GON) follows same pattern as in blood

```{r, message=FALSE, warning=FALSE, cache=TRUE, echo = F, fig.width= 5}
ggdraw() + draw_image("../figures/characterization/prolactin-1.png") 
```

## Fig 4: Hundreds of genes show similar pattern in expression as circulating prolactin

```{r, message=FALSE, warning=FALSE, cache=TRUE, echo = F, fig.width= 5}
ggdraw() + draw_image("../figures/sexes/wgcna-7.png")
```

### 2019-09-06 docs/characterization_manuscript.tex
now with char figs

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Peaks and valleys of prolatin-related gene expression during pigeon
parental care stages} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Rayna M Harris\textsuperscript{\getAff{UC Davis}}\textsuperscript{*},
Suzanne Austin\textsuperscript{\getAff{UC Davis}},
Andrew Lang\textsuperscript{\getAff{UNH}},
Matthew MacManes\textsuperscript{\getAff{UNH}},
Rebecca M Calisi\textsuperscript{\getAff{UC Davis}}\\
\bigskip
\textbf{\getAff{UC Davis}}UC Davis, Davis, CA\\
\textbf{\getAff{UNH}}UNH, smalltown, NH\\
\bigskip
* Corresponding author: rmharris@ucdavis\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The goal of this research was to provide the most comprehensive gene
expression profile of gene expression activity in the HPG axis of male
and female rock doves to provide a deeper undertand of how the
reproductive axis response to typical behavioral transitions associated
with parental care. Here we used RNA sequencing to measure gene activity
at 8 stages of the parental care cyle from nest building, to egg
incubation, and through nestling care. Non-parental groups were added as
controls, but gene expressions differences were assesed between and
across all time points. The three tissue have unique signatures, but
within each tissue we find minial sex differences. The pitutary samples
displacy the msot plasticity in gene expression, with many changes in
gene expression mirroring the typical rise and fall of circulating
prolatin that peaks when chicks hatch. Our analysis provides new insight
into how suites of genes respond in concert to the demands of offspring
care.

@@

Rock doves are socially monogamous and offer bi-parental care, making
inter- and intra-sexual comparisons possible. Birds offer two important
behavioral transition points into parental care: the incubation of eggs
and the caring for chicks. This produces two unique opportunities to
study how the brain transitions into two different suites of parental
care behaviors. Additionally, rock doves exhibit a parental care
strategy analogous to mammals in that they, too, `lactate' to feed their
young (Gillespie et al.~2011, 2012). This lactation, unlike simple
regurgitation of food, consists of the production and sloughing off of
skin cells inside the crop sac of females and males, creating a
protein-rich milk-like substance on which they rear their chicks. Many
functional similarities between rock dove and mammalian lactation exist
concerning the mediation of this event by the hormone prolactin (Dumont
1965). Additionally, like mammalian milk, rock dove milk delivers
essential immunoglobulins and nutritional benefits to young, aiding in
their immune function and development of microbiota {[}2{]}. Thus,
because rock doves incubate eggs and exhibit mammalian-like mediation
and function of lactation for young, they have the potential to serve as
a powerful theoretical bridge to understand the neurobiology of both
avian and mammalian transitions into parental care.

\hypertarget{materials-and-methods-and-results}{%
\section{Materials and Methods and
Results}\label{materials-and-methods-and-results}}

@@

\hypertarget{characterize-changes-in-neural-gene-expression-during-parental-care-transitions}{%
\subsection{Characterize changes in neural-gene expression during
parental care
transitions}\label{characterize-changes-in-neural-gene-expression-during-parental-care-transitions}}

Our working hypothesis is that distinct changes in transcription occur
in the brain at the anticipation of, during, and in response to two
different types of parental care: incubation behavior and hatchling
care.

@@

\textbf{Fig 2. The magnitude of gene expression changes between each
parental transition}. 4-5K genes are differentially expressed beteen
control birds and their nest buliding conspecifics in all tissues except
the male gonad. 500 - 1000 genes are differentially expressed in the
female pituitary from mid-late incubation as well as in the female
hypothalamus and pituitary from hatch to nestingly care day 5.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Peaks and valleys of prolatin-related gene expression during pigeon
parental care stages} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Rayna M Harris\textsuperscript{\getAff{UC Davis}}\textsuperscript{*},
Suzanne Austin\textsuperscript{\getAff{UC Davis}},
Andrew Lang\textsuperscript{\getAff{UNH}},
Matthew MacManes\textsuperscript{\getAff{UNH}},
Rebecca M Calisi\textsuperscript{\getAff{UC Davis}}\\
\bigskip
\textbf{\getAff{UC Davis}}UC Davis, Davis, CA\\
\textbf{\getAff{UNH}}UNH, smalltown, NH\\
\bigskip
* Corresponding author: rmharris@ucdavis\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
The goal of this research was to provide the most comprehensive gene
expression profile of gene expression activity in the HPG axis of male
and female rock doves to provide a deeper undertand of how the
reproductive axis response to typical behavioral transitions associated
with parental care. Here we used RNA sequencing to measure gene activity
at 8 stages of the parental care cyle from nest building, to egg
incubation, and through nestling care. Non-parental groups were added as
controls, but gene expressions differences were assesed between and
across all time points. The three tissue have unique signatures, but
within each tissue we find minial sex differences. The pitutary samples
displacy the msot plasticity in gene expression, with many changes in
gene expression mirroring the typical rise and fall of circulating
prolatin that peaks when chicks hatch. Our analysis provides new insight
into how suites of genes respond in concert to the demands of offspring
care. This data can be used to develop and test hypotheses about the
mechanism regulating parental care behavior.

@@

Rock doves are socially monogamous and offer bi-parental care, making
inter- and intra-sexual comparisons possible. Birds offer two important
behavioral transition points into parental care: the incubation of eggs
and the caring for chicks. This produces two unique opportunities to
study how the brain transitions into two different suites of parental
care behaviors. Additionally, rock doves exhibit a parental care
strategy analogous to mammals in that they, too, `lactate' to feed their
young (Gillespie et al.~2011, 2012). This lactation, unlike simple
regurgitation of food, consists of the production and sloughing off of
skin cells inside the crop sac of females and males, creating a
protein-rich milk-like substance on which they rear their chicks. Many
functional similarities between rock dove and mammalian lactation exist
concerning the mediation of this event by the hormone prolactin (Dumont
1965). Additionally, like mammalian milk, rock dove milk delivers
essential immunoglobulins and nutritional benefits to young, aiding in
their immune function and development of microbiota {[}2{]}. Thus,
because rock doves incubate eggs and exhibit mammalian-like mediation
and function of lactation for young, they have the potential to serve as
a powerful theoretical bridge to understand the neurobiology of both
avian and mammalian transitions into parental care. Our working
hypothesis is that distinct changes in transcription occur in the brain
at the anticipation of, during, and in response to two different types
of parental care: incubation behavior and hatchling care.

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{table-1-comparison-of-total-degs-determined-by-edger-and-deseq2}{%
\subsection{Table 1: comparison of total DEGs determined by edgeR and
DESeq2}\label{table-1-comparison-of-total-degs-determined-by-edger-and-deseq2}}

\hypertarget{fig-1-total-number-of-gene-expression-differences-between-each-stage}{%
\subsection{Fig 1: Total number of gene expression differences between
each
stage}\label{fig-1-total-number-of-gene-expression-differences-between-each-stage}}

@@

\textbf{Fig x. The magnitude of gene expression changes between each
parental transition}. 4-5K genes are differentially expressed beteen
control birds and their nest buliding conspecifics in all tissues except
the male gonad. 500 - 1000 genes are differentially expressed in the
female pituitary from mid-late incubation as well as in the female
hypothalamus and pituitary from hatch to nestingly care day 5.

\hypertarget{fig-2-ability-to-use-gene-expression-to-distingush-groups}{%
\subsection{Fig 2: Ability to use gene expression to distingush
groups}\label{fig-2-ability-to-use-gene-expression-to-distingush-groups}}

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-5-1.pdf}

\hypertarget{fig-3-prolactin-expression-in-pit-but-not-hyp-or-gon-follows-same-pattern-as-in-blood}{%
\subsection{Fig 3: Prolactin expression in PIT (but not HYP or GON)
follows same pattern as in
blood}\label{fig-3-prolactin-expression-in-pit-but-not-hyp-or-gon-follows-same-pattern-as-in-blood}}

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-6-1.pdf}

\hypertarget{fig-4-hundreds-of-genes-show-similar-pattern-in-expression-as-circulating-prolactin}{%
\subsection{Fig 4: Hundreds of genes show similar pattern in expression
as circulating
prolactin}\label{fig-4-hundreds-of-genes-show-similar-pattern-in-expression-as-circulating-prolactin}}

\includegraphics{characterization_manuscript_files/figure-latex/unnamed-chunk-7-1.pdf}
